NMRlipids III: Lipid-cholesterol interactions in atomistic resolution molecular dynamics simulations

### 2017-03-30 MANUSCRIPT/manuscript.tex
New version of manuscript

\title{Quantitative quality of lipid-cholesterol interactions in atomistic resolution molecular dynamics simulations} %Title of paper

@@

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@aalto.fi}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{Aalto University}

@@

%\label{}
\section{Introduction}
Details of intermolecular interactions determine the phase
behaviour and details lateral organization of lipid and bilayers
containing cholesterol \cite{??}.
Formation lateral heterogeneities \cite{??}, lipid rafts \cite{??}
and superlattices \cite{??} in cellular membranes have been
suggested to be driven by interactions between cholesterol
and lipids. While detailed experimental information of these interactions
is relatively sparse, atomistic resolution molecular dynamics simulations
have been widely applied to give detailed explanations of lipid bilayer
lateral organization and lipid-cholesterol interactions \cite{??}.

Simulations qualitatively reproduce the cholesterol condensation effect,
but quantitatve comparison to robust experimental data has not been
typically done. This is partly due to the lack of availability of
systematic experimental data sets for the comparison and partly
due to the lack of protocols for validating intermolecular interactions
in lipid bilayer simulations against experiments.
Scattering experiments are more difficult to interpret for
mixed lipid bilayers than for single component bilayers \cite{??},
thus area per molecule values, one of the main quantity used to
compare simulations to experiments, have not been
available for lipid cholesterol mixtures. Systematic experimental
data set for lipid C-H bond order parameters with different cholesterol
concentrations in POPC bilayer has been published only relatively recently \cite{ferreira13}.

In this work we present also the experimental scattering form factor
data for POPC-cholesterol mixtures by systematically increasing the
cholesterol concentration. Our goal is to show that the combination
of systematically measured C-H bond order parameter and scattering form
factor data can be used to validate the quality of lipid-cholesterol
intermolecular interactions MD simulations. MD simulations can be also
potentially used to give structural interpretation for form factor
measured from mixed systems, which is a major challenge for current methods \cite{??}. 
The approach should be also applicable for mixed lipid bilayers with
other than lipid-cholesterol mixtures.

@@

\subsection{Structural sampling of molecules in bilayers with cholesterol}
Order parameters for C-H bond vectors in lipid bilayer systems, measured
with 13C or 2H NMR techniques, give indirect information about structural
sampling of indidual molecules \cite{ollila16}. The observed increase of
acyl chain order parameters with added cholesterol in simulations and
experiments can be explained by increased trans conformations in acyl chains \cite{??},
which is suggested to play critical role in phase behaviour in PC lipid-cholesterol
mixtures \cite{??}. Consequently, the correct cholesterol ordering effect is
expected to be necessary condition for a model used to understand lipid-cholesterol
phase behaviour.

Acyl chain order parameters for pure 1-palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
and mixture with 50 mol\% of cholesterol from different simulations and experiments
are shown in Fig. \ref{OrderParametersCHOL}. Phase separation is not observed for this system \cite{ferreira13,??},
thus the intermolecular interactions are expected to have maximum effect in equimolar mixture.
Experimental acyl chain order parameters are typically well reproduced in state of the art lipid models
for pure lipid bilayers (for review see \cite{ollila16}). This is also mainly observed in the data
presented here, except that CHARMM36 slightly overestimates the order parameters and MACROG model
underestimates order parameters in the beginning of sn-1 chain. In equimolar mixtures with cholesterol
all models overestimate the order parameters. For Berger/Holtje and MACROG models the overestimation is significant,
while less severe for CHARMM36 and Slipid models.\\
\todo{Is overestimation in CHARMM36 and Slipid significant or not?} \\
\todo{Lipid14 results from recent article should be added in this figure} 
\begin{figure*}[]
  \centering
  \includegraphics[width=17.2cm]{../FIGS/OrderParametersCHOL.eps}
  \caption{\label{OrderParametersCHOL}
    Order parameters from simulations and experiments for acyl chains of  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).
  }
\end{figure*}

@@

\subsection{Lipid bilayer dimensions and density profiles as a function of cholesterol}
Cholesterol induced changes in lipid packing, thickness and density profiles are suggested
to be relevant, for example, for membrane protein interactions and cellular membrane
lateral organization \cite{??}. Experimental and simulation studies show that
the cholesterol induces the so called ''condensing effect'', i.e. acyl chain ordering
increases the membrane thickness and
reduces the area per molecule in lipid bilayers \cite{??}. This is demonstrated
from simulations in Fig. \ref{apls}, showing the area per PC headgroups and per
total number of molecules as a function of cholesterol concentration from different
models. All models show reduction of area per total amount of molecules, partly
due to the smaller area covered by the cholesterol than lipids, but also because
lipids ordered by cholesterol require less space. Due to the latter effect,
the area per PC headgroup do not essentially increase up to the addition of
$\sim$15 mol\% of cholesterol. \\
\todo{Membrane thickness to be plotted as well.}
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/apls.eps}
  \caption{\label{apls}
    Area per molecules calculated from different simulation models
    as a function of cholesterol concentration.
  }
\end{figure}

Bilayer dimensions are indirectly experimentally accessible by scattering methods 
and have been widely used to validate lipid bilayer simulations \cite{??}. 
While sophisticated models are applied to extract area per molecule and thicknes for single
component lipid bilayers, the results for multicomponent systems are more difficult to interpret \cite{??}.
In such case the validation can be done by comparing directly measurable form factors between
simulations and experiments \cite{??}. If simulation reproduces the form factor, it can be also
used to interpret the experiments.  The form factors from different simulation models with
different cholesterol concentrations are compared to experimental data Fig.~\ref{FormFactors}. \\
\todo{Discussion to be finished when the data is solid.}

@@

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
\end{acknowledgments}
\newpage
\appendix
\begin{center}
{\bf SUPPLEMENTARY INFORMATION}
\end{center}
\section{CHARMM36 results from different simulation packages}
The results from CHARMM36 model for lipid bilayers from different 
simulation packages have been reported to give different results in
the literature \cite{piggot12,lee16}. The results are mainly
dependent on different Lennart-Jones cut--off settings, but
all the details are not quite understood. In this work we use
the results from Gromacs 5 with settings suggested to be optimal
by Gromacs webpage. We also compared the results from Gromacs 5 with
these settings to the results simulated with NAMD, OpenMM and literature
values. The comparison is shown in Fig. \ref{programsCOMP}
 \begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/OrderParametersPROGRAMS.eps}

>>>>

\title{NMRlipids III: Lipid-cholesterol interactions in atomistic resolution molecular dynamics simulations} %Title of paper

@@

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

@@

%\label{}
\section{Introduction}
Details of intermolecular interactions determine the phase
behaviour and details lateral organization of lipid and bilayers
containing cholesterol \cite{ipsen87}.
Formation lateral heterogeneities \cite{kinnunen91}, lipid rafts \cite{simons97}
and superlattices \cite{somerharju09} in cellular membranes have been
suggested to be driven by interactions between cholesterol
and lipids. While detailed experimental information of these interactions
is relatively sparse, atomistic resolution molecular dynamics simulations
have been widely applied to give detailed explanations of lipid bilayer
lateral organization and lipid-cholesterol interactions \cite{rog14,sodt14,??}.
However, the simulations must reproduce the measured details, like NMR
order parameters and scattering form factors, to be useful in intepretation
of molecular details in lipid bilayer mixtures.

Simulations qualitatively reproduce the cholesterol condensation effect,
but quantitatve comparison to robust experimental data has not been
typically done. This is partly due to the lack of availability of
systematic experimental data sets for the comparison and partly
due to the lack of protocols for validating intermolecular interactions
in lipid bilayer simulations against experiments.
Scattering experiments are more difficult to interpret for
mixed lipid bilayers than for single component bilayers \cite{pan12,Heftberger15,Marquardt15,??},
thus area per molecule values, one of the main quantity used to
compare simulations to experiments, have not been
available for lipid cholesterol mixtures. Systematic experimental
data set for lipid C-H bond order parameters with different cholesterol
concentrations in POPC bilayer has been published only relatively recently \cite{ferreira13}.

In this work we present also the experimental scattering form factor
data for POPC-cholesterol mixtures by systematically increasing the
cholesterol concentration. Our goal is to show that the combination
of systematically measured C-H bond order parameter and scattering form
factor data can be used to validate the quality of lipid-cholesterol
intermolecular interactions MD simulations. MD simulations can be also
potentially used to give structural interpretation for form factor
measured from mixed systems, which is a major challenge for current methods \cite{pan12,Heftberger15,Marquardt15,??}. 
The approach should be also applicable for mixed lipid bilayers with
other than lipid-cholesterol mixtures.

@@

\subsection{Structural sampling of molecules in bilayers with cholesterol}
Order parameters for C-H bond vectors in lipid bilayer systems, measured
with $^{13}$C or $^{2}$H NMR techniques, give indirect information about structural
sampling of indidual molecules \cite{ollila16}. The observed increase of
acyl chain order parameters with added cholesterol in simulations and
experiments can be explained by increased trans conformations in acyl chains \cite{ferreira13,??},
which is suggested to play critical role in phase behaviour in PC lipid-cholesterol
mixtures \cite{ipsen87}. Consequently, the correct cholesterol ordering effect is
expected to be necessary condition for a model used to understand lipid-cholesterol
phase behaviour.

Acyl chain order parameters for pure 1-palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
and mixture with 50 mol\% of cholesterol from different simulations and experiments
are shown in Fig. \ref{OrderParametersCHOL}. Phase separation is not observed for this system \cite{ionova12,ferreira13},
thus the intermolecular interactions are expected to have maximum effect in equimolar mixture.
Experimental acyl chain order parameters are typically well reproduced in state of the art lipid models
for pure lipid bilayers (for review see \cite{ollila16}). This is also mainly observed in the data
presented here, except that CHARMM36 slightly overestimates the order parameters and MACROG model
underestimates order parameters in the beginning of sn-1 chain. In equimolar mixtures with cholesterol
all models overestimate the order parameters. For Berger/Holtje and MACROG models the overestimation is significant,
while less severe for CHARMM36 and Slipid models.\\
\todo{Is overestimation in CHARMM36 and Slipid significant or not?} \\
\todo{Lipid14 results from recent article should be added in this figure} 
\begin{figure*}[]
  \centering
  \includegraphics[width=17.2cm]{../FIGS/OrderParametersCHOL.eps}
  \caption{\label{OrderParametersCHOL}
    Order parameters from simulations and experiments for acyl chains of  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).
  }
\end{figure*}

@@

\subsection{Lipid bilayer dimensions and density profiles as a function of cholesterol}
Cholesterol induced changes in lipid packing, thickness and density profiles are suggested
to be relevant, for example, for membrane protein interactions and cellular membrane
lateral organization \cite{??}. Experimental and simulation studies show that
the cholesterol induces the so called ''condensing effect'', i.e. acyl chain ordering
increases the membrane thickness and
reduces the area per molecule in lipid bilayers \cite{??}. This is demonstrated
from simulations in Fig. \ref{apls}, showing the area per PC headgroups and per
total number of molecules as a function of cholesterol concentration from different
models. All models show reduction of area per total amount of molecules, partly
due to the smaller area covered by the cholesterol than lipids, but also because
lipids ordered by cholesterol require less space. Due to the latter effect,
the area per PC headgroup do not essentially increase up to the addition of
$\sim$15 mol\% of cholesterol. \\
\todo{Membrane thickness to be plotted as well.}
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/apls.eps}
  \caption{\label{apls}
    Area per molecules calculated from different simulation models
    as a function of cholesterol concentration. Solid marks are area per
    total amount of molecules (chol+PC) and empty marks are area per PC headgroups.
    Top figure shows absolute values and bottom figure shows changes respect
    to pure lipid system.
  }
\end{figure}

Bilayer dimensions are indirectly experimentally accessible by scattering methods 
and have been widely used to validate lipid bilayer simulations \cite{ollila16,??}. 
While sophisticated models are applied to extract area per molecule and thicknes for single
component lipid bilayers, the results for multicomponent systems are more difficult to interpret \cite{pan12,Heftberger15,Marquardt15,??}.
In such case the validation can be done by comparing directly measurable form factors between
simulations and experiments. If simulation reproduces the form factor, it can be also
used to interpret the experiments.  The form factors from different simulation models with
different cholesterol concentrations are compared to experimental data Fig.~\ref{FormFactors}. \\
\todo{Discussion to be finished when the data is solid.}

@@

Cholesterol ordering effect is overestimated in Berger/Holtje and MacRog models.
Slight overestimation is observed also in CHARMM36 and Slipid models, but more
careful analysis is required to conclude if this is significant or not.

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
\end{acknowledgments}
\newpage
\appendix
\begin{center}
{\bf SUPPLEMENTARY INFORMATION}
\end{center}
\section{CHARMM36 results from different simulation packages}
The results from CHARMM36 model for lipid bilayers from different 
simulation packages have been reported to give different results in
the literature \cite{piggot12,lee16}. The results are mainly
dependent on different Lennart-Jones cut--off settings, but
all the details are not quite understood. In this work we use
the results from Gromacs 5 with settings suggested to be optimal
by Gromacs webpage. We also compared the results from Gromacs 5 with
these settings to the results simulated with NAMD, OpenMM and literature
values. Based on comparison shown in Fig. \ref{programsCOMP}, we conclude
that Gromacs 5 with settings suggested in webpage gives consistent
results with the literature and other simulation packages. Thus these
results are used in the main body of the paper. However, order parameters
are slightly overestimated respected to the experiments also with these
settings. \\
\todo{Should be extend this discussion based on discussion at  https://github.com/NMRLipids/NmrLipidsCholXray/issues/4 ?}
 \begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/OrderParametersPROGRAMS.eps}
